FEM Transition Models

### 2023-03-17 FEM_R/Loneliness_Social_Isolation/PAPER_WORKING_SCRIPT.Rmd
Added cognitive function variables and updated lots of transition models. Also added lots to the R notebook and utilities

require(tidyverse)
require(reactable)
require(haven)
require(ggplot2)
require(jtools)
require(ggfortify)
require(survival)
require(survminer)
require(lazyeval)

@@

```{r}
source('../HAZARDS/source/hazard_utils.R')
source('FEM_R/utilities.R')
```

@@

  print(int.name)
  print(paste('Cohort Average Lifeyears:        ', signif(c.avg.lfyr, digits=4)))
  print(paste('Intervention Average Lifeyears:  ', signif(i.avg.lfyr, digits=4)))
  print(paste('Increase:                        ', signif(x = (i.avg.lfyr - c.avg.lfyr), digits=4)))
}

@@

# Compare disability life years between scenarios
disabilityLifeYearComp <- function(c, i, int.name) {
  c.avg.dis <- disabilityLifeYearsCohort(c)
  i.avg.dis <- disabilityLifeYearsCohort(i)
  print(int.name)
  print(paste('Cohort Average Disabled Lifeyears:        ', signif(c.avg.dis, digits=4)))
  print(paste('Intervention Average Disabled Lifeyears:  ', signif(i.avg.dis, digits=4)))
  print(paste('Increase:                        ', signif(x = (i.avg.dis - c.avg.dis), digits=4)))
}

@@

# Compare disability free life years between scenarios
disabilityFreeLifeYearsComp <- function(c, i, int.name) {
  c.avg.dis <- disabilityFreeLifeYearsCohort(c)
  i.avg.dis <- disabilityFreeLifeYearsCohort(i)
  print(int.name)
  print(paste('Cohort Average Disability Free Lifeyears:        ', signif(c.avg.dis, digits=4)))
  print(paste('Intervention Average Disability Free Lifeyears:  ', signif(i.avg.dis, digits=4)))
  print(paste('Increase:                                        ', signif(x = (i.avg.dis - c.avg.dis), digits=4)))
}
```

@@

# Compare disease free life years between scenarios
diseaseFreeLifeYearsComp <- function(c, i, int.name) {
  c.avg.dis <- diseaseFreeLifeYearsCohort(c)
  i.avg.dis <- diseaseFreeLifeYearsCohort(i)
  print(int.name)
  print(paste('Cohort Average Disease Free Lifeyears:        ', signif(c.avg.dis, digits=4)))
  print(paste('Intervention Average Disease Free Lifeyears:  ', signif(i.avg.dis, digits=4)))
  print(paste('Increase:                                        ', signif(x = (i.avg.dis - c.avg.dis), digits=4)))
}
```

@@

#### Disease Free Life Years
```{r}
diseaseFreeLifeYearsComp(cohort_full, no.lonely, 'No Loneliness')
diseaseFreeLifeYearsComp(cohort_full, no.sociso, 'No Social Isolation')
diseaseFreeLifeYearsComp(cohort_full, no.lnly.sociso, 'No Loneliness OR Social Isolation')
```

>>>>

require(tidyverse)
require(haven)
require(ggplot2)
require(jtools)
require(ggfortify)
require(survival)
require(survminer)
require(lazyeval)

@@

```{r}
#source('../HAZARDS/source/hazard_utils.R')
source('FEM_R/utilities.R')
```

@@

  df <- data.frame(outcome = c('LifeYears', 'LifeYears'),
                   intervention = c(int.name, int.name),
                   scenario = c('cohort', 'intervention'),
                   lifeyear = c(c.avg.lfyr, i.avg.lfyr))

  print(int.name)
  print(paste('Cohort Average Lifeyears:        ', signif(c.avg.lfyr, digits=4)))
  print(paste('Intervention Average Lifeyears:  ', signif(i.avg.lfyr, digits=4)))
  print(paste('Increase:                        ', signif(x = (i.avg.lfyr - c.avg.lfyr), digits=4)))

  return(df)
}

@@

# Compare disability life years between scenarios
disabilityLifeYearComp <- function(c, i, int.name) {
  c.avg.dis <- disabilityLifeYearsCohort(c)
  i.avg.dis <- disabilityLifeYearsCohort(i)
  print(int.name)
  print(paste('Cohort Average Disabled Lifeyears:        ', signif(c.avg.dis, digits=4)))
  print(paste('Intervention Average Disabled Lifeyears:  ', signif(i.avg.dis, digits=4)))
  print(paste('Increase:                        ', signif(x = (i.avg.dis - c.avg.dis), digits=4)))
  return(df)
}

@@

# Compare disability free life years between scenarios
disabilityFreeLifeYearsComp <- function(c, i, int.name) {
  c.avg.dis <- disabilityFreeLifeYearsCohort(c)
  i.avg.dis <- disabilityFreeLifeYearsCohort(i)
  df <- data.frame(outcome = c('DisabilityFree', 'DisabilityFree'),
                   intervention = c(int.name, int.name),
                   scenario = c('cohort', 'intervention'),
                   lifeyear = c(c.avg.dis, i.avg.dis))
  print(int.name)
  print(paste('Cohort Average Disability Free Lifeyears:        ', signif(c.avg.dis, digits=4)))
  print(paste('Intervention Average Disability Free Lifeyears:  ', signif(i.avg.dis, digits=4)))
  print(paste('Increase:                                        ', signif(x = (i.avg.dis - c.avg.dis), digits=4)))
  return(df)
}
```

@@

# Compare disease free life years between scenarios
diseaseFreeLifeYearsComp <- function(c, i, int.name) {
  c.avg.dis <- diseaseFreeLifeYearsCohort(c)
  i.avg.dis <- diseaseFreeLifeYearsCohort(i)
  df <- data.frame(outcome = c('DiseaseFree', 'DiseaseFree'),
                   intervention = c(int.name, int.name),
                   scenario = c('cohort', 'intervention'),
                   lifeyear = c(c.avg.dis, i.avg.dis))
  print(int.name)
  print(paste('Cohort Average Disease Free Lifeyears:        ', signif(c.avg.dis, digits=4)))
  print(paste('Intervention Average Disease Free Lifeyears:  ', signif(i.avg.dis, digits=4)))
  print(paste('Increase:                                        ', signif(x = (i.avg.dis - c.avg.dis), digits=4)))
  return(df)
}
```

@@

[1] "No Loneliness"
[1] "Cohort Average Disability Free Lifeyears:         12.55"
[1] "Intervention Average Disability Free Lifeyears:   13.67"
[1] "Increase:                                         1.121"
[1] "No Social Isolation"
[1] "Cohort Average Disability Free Lifeyears:         12.55"
[1] "Intervention Average Disability Free Lifeyears:   12.63"
[1] "Increase:                                         0.08628"
[1] "No Loneliness OR Social Isolation"
[1] "Cohort Average Disability Free Lifeyears:         12.55"
[1] "Intervention Average Disability Free Lifeyears:   13.76"
[1] "Increase:                                         1.208"

#### Disease Free Life Years
```{r}
diseaseFreeLifeYearsComp(cohort_full, no.lonely, 'No Loneliness')
diseaseFreeLifeYearsComp(cohort_full, no.sociso, 'No Social Isolation')
diseaseFreeLifeYearsComp(cohort_full, no.lnly.sociso, 'No Loneliness OR Social Isolation')
```
[1] "No Loneliness"
[1] "Cohort Average Disease Free Lifeyears:         5.48"
[1] "Intervention Average Disease Free Lifeyears:   5.668"
[1] "Increase:                                         0.1873"
[1] "No Social Isolation"
[1] "Cohort Average Disease Free Lifeyears:         5.48"
[1] "Intervention Average Disease Free Lifeyears:   5.623"
[1] "Increase:                                         0.1427"
[1] "No Loneliness OR Social Isolation"
[1] "Cohort Average Disease Free Lifeyears:         5.48"
[1] "Intervention Average Disease Free Lifeyears:   5.852"
[1] "Increase:                                         0.3715"
#### Visualise

Lets write some functions to visualise this information, so will need to collect them into dataframes first.

ly.lnly <- lifeyearComparison(cohort_full, no.lonely, 'L')
ly.sociso <- lifeyearComparison(cohort_full, no.sociso, 'S')
ly.lnly.sociso <- lifeyearComparison(cohort_full, no.lnly.sociso, 'LS')

ly <- rbind(ly.lnly, ly.sociso)
ly.all <- rbind(ly.lnly, ly.sociso, ly.lnly.sociso)
ly.wide <- ly %>% 
  pivot_wider(names_from = scenario,
                              values_from = lifeyear,
                              names_repair = 'unique')
colnames(ly.wide) <- c('outcome', 'scenario', 'cohort', 'intervention')

disab.lnly <- disabilityFreeLifeYearsComp(cohort_full, no.lonely, 'L')
disab.sociso <- disabilityFreeLifeYearsComp(cohort_full, no.sociso, 'S')
disab.lnly.sociso <- disabilityFreeLifeYearsComp(cohort_full, no.lnly.sociso, 'LS')

disab <- rbind(disab.lnly, disab.sociso)
disab.all <- rbind(disab.lnly, disab.sociso, disab.lnly.sociso)
disab.wide <- disab %>% 
  pivot_wider(names_from = scenario,
                              values_from = lifeyear,
                              names_repair = 'unique')
colnames(disab.wide) <- c('outcome', 'scenario', 'cohort', 'intervention')

disease.lnly <- diseaseFreeLifeYearsComp(cohort_full, no.lonely, 'L')
disease.sociso <- diseaseFreeLifeYearsComp(cohort_full, no.sociso, 'S')
disease.lnly.sociso <- diseaseFreeLifeYearsComp(cohort_full, no.lnly.sociso, 'LS')

disease <- rbind(disease.lnly, disease.sociso)
disease.all <- rbind(disease.lnly, disease.sociso, disease.lnly.sociso)
disease.wide <- disease %>% 
  pivot_wider(names_from = scenario,
                              values_from = lifeyear,
                              names_repair = 'unique')
colnames(disease.wide) <- c('outcome', 'scenario', 'cohort', 'intervention')

lifeyears <- rbind(ly.wide, disab.wide, disease.wide)
lifeyears['change'] <- lifeyears['intervention'] - lifeyears['cohort']
```

Now plot.

```{r}
ggplot(data = filter(lifeyears, outcome == 'LifeYears'),
       aes(x = scenario, y = change, color = scenario, fill = scenario)) +
  geom_col()

ggplot(data = filter(lifeyears, outcome == 'DisabilityFree'),
       aes(x = scenario, y = change, color = scenario, fill = scenario)) +
  geom_col()

ggplot(data = filter(lifeyears, outcome == 'DiseaseFree'),
       aes(x = scenario, y = change, color = scenario, fill = scenario)) +
  geom_col()

combined <- ggplot(data = lifeyears, aes(x = factor(scenario, levels = c('L', 'S', 'LS')), y = change, fill = scenario)) +
  geom_col() +
  facet_wrap(~factor(outcome, levels = c('LifeYears', 'DisabilityFree', 'DiseaseFree'))) +
  labs(x = 'Scenario', y = 'Increase') +
  scale_fill_discrete(labels = c('No Loneliness', 'No Social Isolation'))
  #scale_color_manual(labels = c('No Loneliness', 'No Loneliness or Social Isolation', 'No Social Isolation', values = c('red', 'green', 'blue')))

combined

# Save plot into FEM_R/Loneliness_Social_Isolation/plots/
path <- 'FEM_R/Loneliness_Social_Isolation/plots/'

ggsave(filename = 'ly_dfly_facet_L_S.png',
      plot = combined,
      path = path,
      width = 9,
      height = 6)
```

### Sub-populations

Need to get the detailed output for this, but this will be a really nice thing to include.
We can show the improvements in similar plots by different subgroups. Hopefully we'll see bigger improvements in specific subgroups, perhaps a certain age group or gender. The literature I was reading about this suggests that each intervention should affect both genders differently, so should be interesting.

Lets start by appending detailed output and working on a function to get it into a plottable format.

```{r}
cohort <- detailed_output_append(base.out.dir = 'output/LNLY_SOCISO_TEST/',
                              scenario = 'ELSA_core_cohort')
```

```{r}
coh <- cohort

#base2 <- base %>% group_by(hhidpn, mcrep) %>%
#                  mutate(smoken_init = ())

# generate lifeyear for people who have not died
coh$lifeyear <- 1 - coh$died
# generate the number of repetitions for each person
coh$nreps <- max(coh$mcrep) + 1
# now generate the lifeyear intermediate (lifeyears per person)
coh2 <- coh %>% 
  group_by(hhidpn, year) %>%
  mutate(nreps = n())# %>%
  #ungroup() %>%
  #group_by(hhidpn) %>%
  #summarise(n_LY = sum(lifeyear),
  #       mean_LY = n_LY / nreps,
  #       n_DFLY = sum(nodisease),
  #       mean_DFLY = n_DFLY / nreps,
  #       two_yr_LY = mean_LY * 2,
  #       two_yr_DFLY = mean_DFLY * 2,
  #       weight_intermed = sum(weight),
  #       weight2 = weight_intermed / nreps)

#mean <- coh %>% summarise(LY = mean(mean_LY),
#                          DFLY = mean(mean_DFLY))
```

#### Gender
```{r}

print('Female')

### 2023-03-17 FEM_R/Loneliness_Social_Isolation/transition_models_testing.Rmd
Added cognitive function variables and updated lots of transition models. Also added lots to the R notebook and utilities

```{r}
input_dir <- "input_data/"
trans <- read_dta(paste0(input_dir, 'ELSA_transition.dta'))
long <- read_dta(paste0(input_dir, 'ELSA_long.dta'))
```

@@

probit.mort.min <- glm(formula = form.mort.min, family = binomial(link = 'probit'),
                    data = trans)

@@

probit.mort.ls <- glm(formula = form.mort.ls, family = binomial(link = 'probit'),
                    data = trans)

@@

probit.mort.ls2 <- glm(formula = form.mort.ls2, family = binomial(link = 'probit'),
                    data = trans)

@@

probit.mort.full <- glm(formula = form.mort.full, family = binomial(link = 'probit'),
                    data = trans)

@@

@@

>>>>

```{r}
input_dir <- "input_data/"
trans <- read_dta(paste0(input_dir, 'ELSA_transition.dta'))
long <- read_dta(paste0(input_dir, 'ELSA_long.dta'))

# trans needs to be anything after wave 1
trans <- trans %>% 
  filter(wave > 1)

trans.notDead <- trans %>% 
  filter(died == 0)
```

@@

probit.mort.min <- glm(formula = form.mort.min, family = binomial(link = 'probit'),
                    data = trans2)

@@

probit.mort.ls <- glm(formula = form.mort.ls, family = binomial(link = 'probit'),
                    data = trans2)

@@

probit.mort.ls2 <- glm(formula = form.mort.ls2, family = binomial(link = 'probit'),
                    data = trans2)

@@

probit.mort.full <- glm(formula = form.mort.full, family = binomial(link = 'probit'),
                    data = trans2)

@@

### Test

I like the previous model, lets also do a few test versions.

```{r}
form.mort.test <- 'died ~ male + white + hsless + college + 
                  l2age65l + l2age6574 + l2age75p + 
                  l2lnly1 + l2lnly2 +
                  l2sociso1 + l2sociso2 + l2sociso3 + l2sociso4 + l2sociso5 +
                  l2cancre + l2diabe + l2hearte + l2demene + l2lunge +
                  l2stroke + l2smokev + l2smoken + l2smokef +
                  l2alcfreq1 + l2alcfreq2 + l2alcfreq3 + l2alcfreq5 + 
                  l2alcfreq6 + l2alcfreq7 + l2alcfreq8'

probit.mort.test <- glm(formula = form.mort.test, family = binomial(link = 'probit'),
                    data = trans2)

summary(probit.mort.test)
```

@@

# Functional Limitations

Lets have a go at adlstat and iadlstat.

Data:
```{r}
trans.adlstat <- trans %>%
  dplyr::select(idauniq, wave, died, adlstat, iadlstat, male, white, hsless, 
                college, l2age65l, l2age6574, l2age75p, l2stroke, l2demene, 
                l2alzhe, l2osteoe, l2catracte, l2lnly, l2lnly1, l2lnly2, 
                l2sociso, l2cesd, l2cohab, l2widowed, l2single, l2married,
                l2alcfreq, l2smoken, l2smokef, l2smokev, l2tr20, l2verbf,
                l2orient) %>%
  filter(wave > 1) %>%
  filter(died == 0)
trans.adlstat.complete <- trans.adlstat[complete.cases(trans.adlstat),]

trans.adlstat.complete$adlstat <- as.factor(trans.adlstat.complete$adlstat)
trans.adlstat.complete$iadlstat <- as.factor(trans.adlstat.complete$iadlstat)
```

## ADLSTAT

```{r}
## MINIMAL
form.adlstat.min <- 'adlstat ~ male + scale(l2age65l) + scale(l2age6574) + scale(l2age75p)'
probit.adlstat.min <- polr(formula = form.adlstat.min,
                    data = trans.adlstat.complete,
                    Hess = TRUE,
                    na.action = na.omit)
print('Minimal model:')
summary(probit.adlstat.min)

## FULL
form.adlstat <- 'adlstat ~ male + white + hsless + college + scale(l2age65l) + scale(l2age6574) + scale(l2age75p) + l2stroke + l2demene + l2alzhe + l2osteoe + l2catracte + l2lnly + l2sociso'
probit.adlstat <- polr(formula = form.adlstat,
                    data = trans.adlstat.complete,
                    Hess = TRUE,
                    na.action = na.omit)
print('Full model:')
summary(probit.adlstat)
```

Now testing.

```{r}
## TEST
form.adlstat.test <- 'adlstat ~ male + white + hsless + college + 
scale(l2age65l) + scale(l2age6574) + scale(l2age75p) + l2stroke + l2demene + l2osteoe + l2catracte + l2lnly + l2sociso + l2cesd + l2married + 
l2widowed + l2alcfreq + l2smokev + l2tr20 + l2verbf + l2orient'
probit.adlstat.test <- polr(formula = form.adlstat.test,
                    data = trans.adlstat.complete,
                    Hess = TRUE,
                    na.action = na.omit)
print('Test model:')
summary(probit.adlstat.test)
```

AIC: 64881.53
AIC: 64857.35 : l2tr20
AIC: 57179.64 : l2verbf
AIC: 57008.17 : l2orient
AIC: 57006.91 : removed l2single

Stay there for now?

## IADLSTAT

```{r}
## MINIMAL
form.iadlstat.min <- 'iadlstat ~ male + scale(l2age65l) + scale(l2age6574) + scale(l2age75p)'
probit.iadlstat.min <- polr(formula = form.iadlstat.min,
                    data = trans.adlstat.complete,
                    Hess = TRUE,
                    na.action = na.omit)
print('Minimal model:')
summary(probit.iadlstat.min)

## FULL
form.iadlstat <- 'iadlstat ~ male + white + hsless + college + scale(l2age65l) + scale(l2age6574) + scale(l2age75p) + l2stroke + l2demene + l2alzhe + l2osteoe + l2catracte + l2lnly + l2sociso'
probit.iadlstat <- polr(formula = form.iadlstat,
                    data = trans.adlstat.complete,
                    Hess = TRUE,
                    na.action = na.omit)
print('Full model:')
summary(probit.iadlstat)
```

Now test.

```{r}
## TEST
form.iadlstat.test <- 'iadlstat ~ male + white + hsless + college + 
scale(l2age65l) + scale(l2age6574) + scale(l2age75p) + l2stroke + l2demene + l2osteoe + l2catracte + l2lnly + l2sociso + l2cesd + l2married + 
l2widowed + l2alcfreq + l2smokev + l2tr20 + l2verbf + l2orient'
probit.iadlstat.test <- polr(formula = form.iadlstat.test,
                    data = trans.adlstat.complete,
                    Hess = TRUE,
                    na.action = na.omit)
print('Test model:')
summary(probit.iadlstat.test)
```

# Cognitive Function Vars

## TR20

```{r}
## MINIMAL
form.tr20.min <- 'tr20 ~ male + scale(l2age65l) + scale(l2age6574) + scale(l2age75p)'
probit.tr20.min <- lm(formula = form.tr20.min,
                    data = trans.notDead,
                    na.action = na.omit)
print('Minimal model:')
summary(probit.tr20.min)

## TEST
form.tr20 <- 'tr20 ~ male + white + hsless + college + scale(l2age65l) + scale(l2age6574) + scale(l2age75p)'
probit.tr20 <- lm(formula = form.tr20,
                    data = trans.notDead,
                    na.action = na.omit)
print('Full model:')
summary(probit.tr20)
```

```{r}
## TEST
form.tr20.test <- 'tr20 ~ male + white + hsless + college + scale(l2age65l) + scale(l2age6574) + scale(l2age75p) + l2hearing + l2sight + l2adlstat + l2iadlstat + l2cesd + l2smoken + l2employed + l2retired'
probit.tr20.test <- lm(formula = form.tr20.test,
                    data = trans.notDead,
                    na.action = na.omit)
print('Full model:')
summary(probit.tr20.test)

# l2lnly
##  + l2physact + l2hearing + l2adlstat + l2iadlstat + l2socyr + l2cesd + l2smoken + l2smoke
```
